---
title: "Variance Simulator"
author: "Gert"
date: "22 April 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Idea

The Idea behind the application is to be able to simulate rolls of dice and 
calculate their variance. There are two parameters that can be changed for 
the simulation.

* Number of Simulations
* Number of Dice

## Theory

In theory the variance of the simulated dice rolls should, when executed
an infinite number of times, converge to 105/36 = 2.92.

Because the expected value of dice rolls is 3.5 we can calculate the expected
variance with: 

(1−3.5)^2 + (2−3.5)^2 + (3−3.5)^2 + (4−3.5)^2 + (5−3.5)^2 + (6−3.5)^2 = 105/36


## Calculation

When calculating the sample variance distribution with parameters:

* Number of Simulations = 1000
* Number of Dice = 10

```{r echo = TRUE}
df <- data.frame(
    variance = c(apply(matrix(sample(1 : 6, 1000 * 10, replace = TRUE), 
                       1000), 1, var)))
mean <- mean(df$variance)
```

we get the following result for the mean sample variance `r mean`.

## Plot of Variance Distribution from the simulated Data

```{r}
library(ggplot2)
ggplot(df, aes(x = variance)) + 
     geom_histogram(alpha = .20, binwidth=.3, colour = "black") +
     geom_vline(xintercept = mean(df$variance), size = 2)
```
